Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-08 app-trees.tex
appendices et commentaires

% SWT without epsilon-transitions
% SWT do not read the other tape in "skip" transitions 10 or 01
%     (hence this version is simpler  than V2)
% SWA without epsilon-transitions
% SW-VPA read top of stack (when not empty) at every transition
%        (6 kinds of transitions).

% default style article.cls
%\input{main-article}

% Springer llncs style
% https://www.springer.com/gp/computer-science/lncs/conference-proceedings-guidelines
%\input{main-llncs}

% EPTCS style
% http://www.eptcs.org
%\input{main-eptcs}

% LIPICS v.2021 style
% https://submission.dagstuhl.de/documentation/authors
\input{main-lipics}

% comments in margin with \todo command
\usepackage[colorinlistoftodos]{todonotes} % option disable to hide todos
% command todo : options
% - noline: no line connecting the note with the place in the text where the note occurs in the latex code
% - fancyline : curved arrow from note to inssertion point
% - inline: place a todonote inside the text instead of in the margin
% - author=name
\newcommand{\florent}[1]{\todo[noline,size=\tiny,color=yellow!40]{#1}}
\newcommand{\lydia}[1]{\todo[noline,size=\tiny,color=red!30]{#1}}
\newcommand{\philippe}[1]{\todo[noline,size=\tiny,color=green!30]{#1}}

\usepackage{hyperref}
%\usepackage[bookmarks,bookmarksnumbered,naturalnames,plainpages=false]{hyperref}
%usepackage{url}

% for footnote ref
\usepackage{refcount}

% array and tabular
\usepackage{array}
\newcolumntype{C}[1]{>{\centering\arraybackslash\hspace{0pt}}p{#1}}

% extension of enumerate env. (style for displaying counters)
% \usepackage{enumerate}

%% pictures
% \usepackage{graphicx}
% \DeclareGraphicsExtensions{.pdf,.png,.jpg}
% \graphicspath{fig/}

%% PGF, Tikz
\usepackage{tikz-cd}
%% \usepackage{pgfplots}
%% \usepgfplotslibrary{dateplot}
%% \usepackage{pgf,pgfarrows,pgfnodes, pgfautomata}
% \usepackage{tikz}
% \usetikzlibrary{cd}
%% \usetikzlibrary{arrows}
%% \usetikzlibrary{calc}
%% \usetikzlibrary{snakes}
%% \usetikzlibrary{backgrounds}
% \usetikzlibrary{trees}
%% \usetikzlibrary{automata}
%% \usetikzlibrary{positioning}
%% \usetikzlibrary{matrix}
%% \usetikzlibrary{patterns}
%% \usetikzlibrary{shapes}

% symbols
\usepackage{amsmath}
\allowdisplaybreaks
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{bbold}
\usepackage{latexsym}
%\usepackage{amsfonts}
\usepackage{stmaryrd}
%\usepackage{mathabx}
%\usepackage{MnSymbol}
\usepackage{harmony} % simple music fonts
\usepackage{mathtools} % for arrows
%\usepackage{mathptmx}
\usepackage{comment}

%% algorithms
%\usepackage{algorithm}
%\usepackage{program}
\usepackage[ruled,vlined]{algorithm2e}

%% allows for page break within arrays
\usepackage{longtable}

%% arrows etc
%\input{rewriting}

%% music symbols
% see http://tug.ctan.org/info/latex4musicians/latex4musicians.pdf
\usepackage{musicography}

%% for new macros
\usepackage{xspace}

%% Misc macros
\def\ie{\textit{i.e.}\xspace}
\def\eg{\textit{e.g.}\xspace}
\def\wrt{\textit{wrt}\xspace}
%\def\wlog{\textit{wlog}\xspace}
\def\etc{\textit{etc}\xspace}

\def\<#1>{\langle #1 \rangle}
\newcommand{\pair}[2]{\langle{#1}, {#2}\rangle}

%\newcommand{\config}[2]{\ensuremath \genfrac{[}{]}{0pt}{}{#1}{#2}}
\newcommand{\config}[2]{\ensuremath{#1}[{#2}]}
\newcommand{\configup}[2]{\ensuremath{#1}\left[\begin{array}{c} #2 \end{array}\right]}
\def\stacksep{\cdot}
\def\stackup{\\}

\newcommand{\opluseq}{\ensuremath\mathrel{\oplus}=}

% \bigominus
\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{
<-6> mathx5 <6-7> mathx6 <7-8> matha7
<8-9> mathx8 <9-10> mathx9
<10-12> mathx10 <12-> mathx12
}{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareMathSymbol{\bigominus}{\mathop}{mathx}{"C1}

\newcommand{\A}{\mathcal{A}} % automata, VPA
\newcommand{\B}{\mathcal{B}} % automata (in constructions)
\newcommand{\C}{\mathcal{C}} % automata (in constructions)
\newcommand{\D}{\mathcal{D}}  % transducer = distance
\newcommand{\E}{\mathbb{E}}
%\newcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\W}{\mathbb{W}}
\newcommand{\Semiring}{\mathbb{S}}
\newcommand{\zero}{\mathbb{0}}
\newcommand{\one}{\mathbb{1}}
\newcommand{\dom}{\ensuremath{\mathit{dom}}}

\def\SA{\textsf{sA}\xspace}
\def\WA{\textsf{wA}\xspace}
\def\SW{\textsf{sw}\xspace}
\def\swM{\textsf{swM}\xspace}
\def\SWT{\textsf{swT}\xspace}
\def\SWA{\textsf{swA}\xspace}
\def\SWTA{\textsf{swTA}\xspace}
\def\SWVPA{\textsf{sw-VPA}\xspace}
\def\VPA{\textsf{VPA}\xspace}
\def\SVPA{\textsf{sVPA}\xspace}
\def\weight{\mathsf{weight}}
\def\wei{\mathsf{w}}
\def\mei{\mathsf{m}}
\def\init{\mathsf{in}}
\def\final{\mathsf{out}}
\def\Omegai{{\Omega_\mathsf{i}}}
\def\Omegac{{\Omega_\mathsf{c}}}
\def\Omegar{{\Omega_\mathsf{r}}}
\def\Sigmai{{\Sigma_\mathsf{i}}}
\def\Sigmac{{\Sigma_\mathsf{c}}}
\def\Sigmar{{\Sigma_\mathsf{r}}}
\def\Deltai{{\Delta_\mathsf{i}}}
\def\Deltac{{\Delta_\mathsf{c}}}
\def\Deltar{{\Delta_\mathsf{r}}}
\def\Phii{{\Phi_\mathsf{i}}}
\def\Phic{{\Phi_\mathsf{c}}}
\def\Phir{{\Phi_\mathsf{r}}}
\def\Phix{{\Phi_\mathsf{x}}}
\def\Phici{{\Phi_\mathsf{ci}}}
\def\Phicc{{\Phi_\mathsf{cc}}}
\def\Phicr{{\Phi_\mathsf{cr}}}
\def\Phicx{{\Phi_\mathsf{cx}}}
\def\weii{{\wei_\mathsf{i}}}
\def\weic{{\wei_\mathsf{c}}}
\def\weir{{\wei_\mathsf{r}}}
\newcommand{\weie[1]}{\wei_{#1}^\mathsf{e}}
\def\weiei{\weie[\mathsf{i}]}
\def\weiec{\weie[\mathsf{c}]}
\def\weier{\weie[\mathsf{r}]}
\def\weiex{\weie[\mathsf{x}]}
\newcommand{\ioi}[1]{\mathsf{ioi}({#1})}
\newcommand{\rank}{\mathsf{rk}}
\newcommand{\lin}{\mathsf{lin}}

\newcommand{\call}[1]{\ensuremath #1} %{\ensuremath \langle_{#1}}
\newcommand{\return}[1]{\ensuremath #1} %{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$
\newcommand{\ccall}[1]{\ensuremath \langle_{#1}}
\newcommand{\creturn}[1]{\ensuremath {}_{#1}{\rangle}} % $\prescript{}{a}{)}$

%\sloppy

% Parsing over infinite alphabet as optimal alignment computation
% as edit-distance between string and language
%
%\title{Symbolic Weighted Parsing and Automated Music Transcription}
%\title{Symbolic Weighted Language Models and Automated Music Transcription}
\title{Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets}
%\title{Weighted Visibly Pushdown Automata and Automated Music Transcription}
%\titlerunning{WVPA \& AMT}
%\titlerunning{Symbolic Weighted Parsing} % and Automated Music Transcription
%\authorrunning{Florent Jacquemard}

\date{\today}

\begin{document}
\thispagestyle{empty}
\maketitle

\begin{abstract}
\input{abstract}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% intro
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction} \label{sec:intro}
\input{introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% prelin
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Preliminary Notions}
\label{section:prelim}\label{sec:prelim}

%notations: for set $S$ : sets of sequences $S^*$ and $S^+$...
%interval $[i..j]$ of natural numbers

\paragraph*{Semirings}
\label{section:semiring}\label{sec:semiring}
\input{semiring}

\paragraph*{Label Theory}
\label{section:symbols}
\input{labels}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SWT & SWA
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{SW Automata and Transducers}
\label{section:transducer}\label{sec:transducer}
\label{section:SWA}\label{sec:SWA}
\label{section:SWT}\label{sec:SWT}

We follow the approach of~\cite{Mohri03EDWA} for the computation of distances,
between words and languages, using weighted transducers,
and extend it to infinite alphabets.
% with models of symbolic weighted automata and transducers.
%
The models introduced in this section generalize
weighted automata and transducers~\cite{Droste09handbook}
%over finite alphabets, see  e.g.~\cite{Mohri03EDWA},
by labeling each transition with a weight function (instead of a simple weight value),
that takes the input and output symbols as parameters.
These functions are similar to the guards of symbolic automata~\cite{dAntoniVeanes17CAV,dAntoni21CACM},
but they can return values in a generic semiring,
whereas the latter guards are restricted to the Boolean semiring.
%

%\subsection{Definitions} \label{sec:SWTdef}\label{sec:SWAdef}
%(SWT)
\noindent
Let $\Semiring$ be a commutative semiring,
$\Sigma$ and $\Delta$ be alphabets called respectively \emph{input} and \emph{output}, %{alphabets},
and $\bar\Phi$ be a label theory over $\Semiring$
containing $\Phi_\Sigma$, $\Phi_\Delta$, $\Phi_{\Sigma, \Delta}$.

\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over $\Sigma$, $\Delta$, $\Semiring$ and $\bar\Phi$
%the input and output alphabets~$\Sigma$ and $\Delta$ with label theory $\bar\Phi$, and the semiring $\Semiring$
is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states,
$\mathsf{in} : Q \to \Semiring$   %\Phi_{\Sigma, \Delta}
(respectively $\mathsf{out} : Q \to \Semiring$)  %\Phi_{\Sigma, \Delta}
are functions defining the weight for entering
(respectively leaving) computation in a state,
and $\bar{\wei}$ is a triplet of transition functions
$\wei_{10}: Q \times Q \to \Phi_{\Sigma}$,
$\wei_{01}: Q \times Q \to \Phi_{\Delta}$, and
$\wei_{11}: Q \times Q \to \Phi_{\Sigma, \Delta}$.
\end{definition}
%
\noindent
We call \emph{number of transitions} of $T$ the number of pairs of states
$q, q' \in Q$ such that $\wei_{10}$ or $\wei_{01}$ or $\wei_{11}$
is not the constant $\zero$.
%
\noindent
For convenience, we shall sometimes present transitions
as functions of
$Q \times (\Sigma \cup \{ \varepsilon \}) \times (\Delta \cup \{ \varepsilon \}) \times Q \to \Semiring$,
overloading the function names,
such that, for all $q, q' \in Q$, $a \in \Sigma$,  $b \in \Delta$,

\[
\begin{array}{rcll}
\wei_{10}(q, a, \varepsilon, q') & = & \phi(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi = \wei_{10}(q, q') \in \Phi_\Sigma,\\
\wei_{01}(q, \varepsilon, b, q') & = & \psi(b) &
\quad\mathrm{where~} \psi = \wei_{01}(q, q') \in \Phi_\Delta,\\
\wei_{11}(q, a, b, q') & = & \eta(a, b) &
\quad\mathrm{where~} \eta = \wei_{11}(q, q') \in \Phi_{\Sigma, \Delta}.\\
\end{array}
\]
%
\noindent
% The symbolic-weighted transducer
The \SWT $T$ computes on pairs of words $\< s, t> \in \Sigma^* \times \Delta^*$,
$s$ and $t$, being respectively called \emph{input} and \emph{output} word.
% \noindent
More precisely, $T$ defines a mapping
from $\Sigma^* \times \Delta^*$ into~$\Semiring$,
based on an intermediate function $\weight_T$
defined recursively, for every states $q, q' \in Q$,
and every pairs of strings $\< s, t> \in \Sigma^* \times \Delta^*$,
% \setminus \{ \< \varepsilon, \varepsilon >\}$,
where $au$, and $bv$, denote the concatenation
of the symbol $a \in \Sigma$ (resp. $b \in \Delta$)
with a word $u \in \Sigma^*$ (resp. $v \in \Delta^*$).
\lydia{added $u$ and $v$ def}
%
%$a \in \Sigma$, $u \in \Sigma^*$, $b\in \Delta$, $v\in \Delta^*$, by:
%
\begin{align}
%\weight_T(q, \varepsilon, \varepsilon, q)  & = \one \label{eq:SWT-weight}\\ %\final(q)\\
%\weight_T(q, \varepsilon, \varepsilon, q') & = \zero \quad \mathrm{if~} q \neq q'\nonumber\\
\weight_T(q, \varepsilon, \varepsilon, q')  & = \one
 \quad \mathrm{if~} q = q' \mathrm{~and~} \zero \mathrm{~otherwise} \label{eq:SWT-weight}\\
\weight_T(q, s, t, q') & = \displaystyle\bigoplus_{\begin{array}{c}
                                                   \scriptstyle q'' \in Q\\[-1ex]
                                                   \scriptstyle s = au,\, a \in \Sigma
                                                   \end{array}}
    \wei_{10}(q, a, \varepsilon, q'') \otimes \weight_T(q'', u, t, q')\nonumber\\
                    & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                                     \scriptstyle q'' \in Q\\[-1ex]
                                                     \scriptstyle t = bv,\, b \in \Delta\\
                                                     \end{array}}
    \wei_{01}(q, \varepsilon, b, q'') \otimes \weight_T(q'', s, v, q')\nonumber\\
                    & \oplus \displaystyle\bigoplus_{\begin{array}{c}
                                                     \scriptstyle q'' \in Q\\[-1ex]
                                                     \scriptstyle s = au,\, t = bv\\
                                                     \end{array}}
    \wei_{11}(q, a, b, q'') \otimes \weight_T(q'', u, v, q')\nonumber
\end{align}

%
We recall that, by convention (Section~\ref{sec:semiring}),
an empty sum with $\bigoplus$ is equal to~$\zero$.
%
Intuitively, using a transition $\wei_{ij}(q, a, b, q')$ means for $T$:
when reading respectively $a$ and $b$ at the current positions in the input and output words,
increment the current position in the input word if and only if $i = 1$,
and in the output word iff $j = 1$, %(otherwise, do not change it),
and change state from $q$ to $q'$.
When $a = \varepsilon$ (resp. $b = \varepsilon$), the current symbol
in the input (resp. output) is not read.
%
%In contrast with the models of weighted transducers over finite alphabets~\cite{Mohri03ijfcs},
%the input and output symbols at current positions are always read by transitions,
%even when they do not change the reading position the head's position.
%This is an important feature in the case of an infinite alphabet in
%order to compare input and output symbols.
%which cannot be stored in the finite memory of the transducer.
%
Since~$\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei_{ij}(q, a, b, q'')$ equal to $\zero$ in the above expression
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and $b$.
This is analogous to the case of a transition's guard not satisfied by $\<a, b>$ for
symbolic transducers.

%whereas considering the current symbol may be useful to compute a transition weight.
%(even when it does not change the head's position, like with $\varepsilon$-transitions).
%
%The cases $\weight_T(q, au, \varepsilon, q')$ and $\weight_T(q, \varepsilon, bv, q')$
%are missing in the definition of $\weight_T$.
%It means that $T$ must avoid configurations where it reached the end of
%the output word and not of the input one, or vice-versa.
%This can be done by using $\wei_{10}$ and $\wei_{01}$
%before reaching the end of word, and using a special state for this purpose.

The expression \eqref{eq:SWT-weight}
can be seen as a stateful definition of
an edit-distance between a word $s \in \Sigma^*$ and a word $t \in \Delta^*$,
see also~\cite{Mohri03ijfcs}.
Intuitively,
$\wei_{10}(q, a, \varepsilon, r)$ is the cost of
the deletion of the symbol~$a \in \Sigma$ in~$s$,
$\wei_{01}(q, \varepsilon, b, r)$ is the cost
of the insertion of~$b \in \Delta$ in $t$,
and $\wei_{11}(q, a, b, r)$ is the cost
of the substitution of  $a \in \Sigma$ by~$b \in \Delta$.
%
The cost of a sequence of such operations transforming $s$ into $t$,
is the product, with $\otimes$, of the individual costs of the operations involved;
and the distance between $s$ and $t$ is the sum, with $\oplus$,
of all possible products.
%
\medskip\noindent
%Let $\< s, t> \in \Sigma^* \times \Delta^*$, with $s = s_1\ldots s_n$, and $t = t_1\ldots t_m$.
Formally, the weight associated by $T$ to $\< s, t> \in \Sigma^* \times \Delta^*$ is:
%defined as follows:
\begin{equation}
T(s, t)  =
\displaystyle\bigoplus_{q, q' \in Q} \init(q)
\mathop{\otimes} \weight_T(q, s, t, q') \mathop{\otimes} \final(q')
\label{eq:SWT-value}
\end{equation}

%\begin{example}
%We can define with a \SWT the computation of a similarity measure
%between timed sequences
%similar to dynamic time warping (DTW).
%
%Let $\Semiring$  be the tropical (\emph{min-plus}) semiring of Figure~\ref{fig:semirings} and
%let $\Sigma = \Delta = \mathbb{R}_+$ be sets of timestamps.
%We consider a \SWT with one state $q$ and transitions
%$\wei_{11}(q, d, d', q) =
% \wei_{01}(q, d, d', q) =
% \wei_{10}(q, d, d', q) = |d' - d|$,
%for all $d, d' \in \mathbb{R}_+$.
%% needs reading input/output symbols even by epsilon-transitions.
%The recursive definition of $\weight_T$ correspond to the dynamic programming equations of DTW
%for the computation of an optimal match between words,
%the matching cost for two symbols being the
%the time distance between them.
%\endex
%\end{example}

\begin{example}
Let us develop the example of comparison between music played by a performer,
represented as a sequence $s \in \Sigma^*$ of events in the MIDI alphabet $\Sigma$,
and  a music score represented as a sequence $t \in \Delta^*$
in the CWMN alphabet $\Delta$. We build a small weighted transducer model
 with two states $q_0$ and $q_1$ that
calculates the distance between $s$ and $t$.

If one performed event $s_i$  corresponds
to one notated event $t_1$ (for instance MIDI code 61 and pitch A4),
the weight value computed by the \SWT is the time distance between both,
as in Example~\ref{distance-time}, and is modeled by
transitions $\wei_{11}$ below.
%
If we meet the music notation symbol '-' that
represents continuation (such as instance in \emph{ties}
$\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
or \emph{dots} \musQuarterDotted{}), it is  skipped with no cost (transitions $\wei_{01}$ or weight $\one$).
%transitions
\[
\begin{array}{rclcrcl}
\wei_{11}(q_0, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d| & \quad &
\wei_{11}(q_1, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d|\\
\wei_{01}(q_0, \varepsilon, \< \mathsf{-}, d'>, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \< \mathsf{-}, d'>, q_0) & = & \one\\
\wei_{10}(q_0, d, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also must be able to take performing errors into account, while still being able to compare with the score,
since a performer could, for example, play an unwritten extra note.
\lydia{reformulated this sentence}
%
This is modelled by the transition $\wei_{10}$ with an arbitrary weight value $\alpha \in \Semiring$,
switching from state $q_0$ (normal) to $q_1$ (error).
\philippe{Comprends pas cette phrase}
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.

\lydia{ccl to the ex}
That way, an \SWT is capable of evaluating the differences between a score and a performance,
all the while ensuring that performance errors are plausible.
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

\begin{comment}
\begin{example}
In Common Western Music Notation~\cite{Gould11Notation},
several symbols may be used to represent one single sounding event.
\lydia{unique $\to$ similar}
\florent{similar $\to$ single}
For instance, several notes can be combined with a tie,
like in $\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
and one note can be augmented by half its duration with a dot like in~\musQuarterDotted{}.
\lydia{modif.}
These notations are perceived equivalent when played, as their duration is equal, yet the notation is different.
We thus want to be able to compare a music score with music played by a performer.
\lydia{changed end}
%
\noindent
We propose a small weighted transducer model with two states $q_0$ and $q_1$
that calculates the distance bewteen an input sequence of sounding
events (music "performance")
to an output sequence of written events (music "score").
% simple pointwise distance between two sequences of timestamped events **
%
Let us consider the tropical (\emph{min-plus}) semiring~$\Semiring$
of Figure~\ref{fig:semirings} and
let $\Sigma = \mathbb{R}_+$ be an input alphabet of event dates
and $\Delta = \{ \mathsf{e}, \mathsf{-} \} \times \mathbb{R}_+$
be an output alphabet of symbols with timestamps.
A symbol $\< \mathsf{e}, d > \in \Delta$ represents an event starting at date $d$,
and $\< \mathsf{-}, d >$ is a continuation of the previous event.
%More precisely, we let
%$\Sigma = \{ a, - \} \times \R_+$ and
%This example of $\Delta$ is motivated by the case of music notation,
%where several notated events (notes) can be tied together,
%with a \emph{tie} or a \emph{dot}
%meaning that they will be played as a unique sounding event.
%The timestamp of $a \in \Sigma$, denoted by $\mathsf{t}(a)$, is expressed as a rational number.

We consider a \SWT with two states $q_0$ and $q_1$ whose purpose
is to compare a recorded performance $s \in \Sigma^*$
with a notated music sheet $t \in \Delta^*$.
One timestamp $d_i \in \Sigma$ may correspond
to one notated event $\<\mathsf{e}, d'_i> \in \Delta$, in which case
the weight value computed by the \SWT is the time distance between both
(see transitions $\wei_{11}$ below).
%
If $\<\mathsf{e}, d'_i>$ is followed by continuations
$\<\mathsf{-}, d'_{i+1}>$..., they are just skipped with no cost (transitions $\wei_{01}$ or weight $\one$).
%transitions
\[
\begin{array}{rclcrcl}
\wei_{11}(q_0, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d| & \quad &
\wei_{11}(q_1, d, \< \mathsf{e}, d'>, q_0) & = & |d' - d|\\
\wei_{01}(q_0, \varepsilon, \< \mathsf{-}, d'>, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \< \mathsf{-}, d'>, q_0) & = & \one\\
\wei_{10}(q_0, d, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also must be able to take performing errors into account, while still being able to compare with the score,
since a performer could, for example, play an unwritten extra note.
\lydia{reformulated this sentence}
%
This is modelled by the transition $\wei_{10}$ with an arbitrary weight value $\alpha \in \Semiring$,
switching from state $q_0$ (normal) to $q_1$ (error).
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.

\lydia{ccl to the ex}
That way, an \SWT is capable of evaluating the differences between a score and a performance,
all the while ensuring that performance errors are plausible.

%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}
\end{comment}

\noindent
The \emph{Symbolic Weighted Automata} %$A = \< Q, \init, \weight, \final >$
%over $\Sigma$,  $\Semiring$ and $\bar\Phi$
are defined similarly as the transducers of Definition~\ref{def:SWT},
by simply omitting the output symbols.
%
%In this case, the label theory $\bar\Phi$ can be reduced to a singleton $\< \Phi_\Sigma>$.
%
\begin{definition} \label{def:SWA}
A \emph{symbolic-weighted automaton} (\SWA)
over $\Sigma$, $\Semiring$ and $\bar\Phi$
%the input alphabet~$\Sigma$ and the commutative semiring $\Semiring$
is a tuple
$A = \< Q, \init, {\wei_1}, \final >$,
where $Q$ is a finite set of states,
$\mathsf{in} : Q \to \Semiring$ %\Phi_\Sigma$,
(respectively $\mathsf{out} : Q \to \Semiring$)  %\Phi_\Sigma,$
are functions defining the weight for entering
(respectively leaving) computation in a state,
and ${\wei_1}$ is a transition function
from $Q \times Q$ into~$\Phi_{\Sigma}$.
\end{definition}
%
\noindent
As above in the case of \SWT,
when $\wei_1(q, q') = \phi \in \Phi_\Sigma$,
%respectively $\mathsf{in}(q) = \phi$, $\mathsf{out}(q') = \phi$,
we may write $\wei_1(q, a, q')$ for~$\phi(a)$.
%respectively $\mathsf{in}(q, a) = \phi(a)$, $\mathsf{out}(q', a) = \phi(a)$.
%$\wei_1: Q \times \Sigma \times Q \to \Semiring$,
The computation of $A$ on words $s \in \Sigma^*$
is defined with an intermediate function $\weight_A$,
defined as follows for $q, q' \in Q$, $a \in \Sigma$, $u \in \Sigma^*$,
%
\begin{align}
\weight_A(q, \varepsilon, q) & = \one\\ %\final(q)\\
\weight_A(q, \varepsilon, q') & = \zero \quad \mathrm{if~} q \neq q'\nonumber\\
\weight_A(q, au, q') & =  \displaystyle\bigoplus_{q'' \in Q}
    \wei_{1}(q, a, q'') \otimes \weight_A(q'', u, q')\nonumber
\label{eq:SWA-weight}
\end{align}
%
\noindent
and the weight value associated by $A$ to
$s \in \Sigma^*$ is defined as follows: %$s = s_1\ldots s_n \in \Sigma^+$
\begin{equation}
A(s)  =
\displaystyle\bigoplus_{q, q' \in Q} \init(q)
\mathop{\otimes} \weight_A(q, s, q') \mathop{\otimes} \final(q')
\label{eq:SWA-value}
\end{equation}

%
%When $\wei_\varepsilon(q, q') = \zero$ for all $q, q' \in Q$,
%the automaton~$A$ is called \emph{without $\varepsilon$-transitions}.

%The \emph{summary} of a $\SWT$, resp. a $\SWA$, is ***

%\subsection{Properties}
\noindent
\philippe{Il me manque une explication:  on construit un automate
qui, étant donnée une partition $t$, renvoie la distance minimale
avec n'importe quelle performance (distance donnée par un transducer)?
Quel est le rôle de A(s)?
}
The following property will be useful to the approach on
symbolic weighted parsing presented in Section~\ref{sec:parsing}.

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$,
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and a $\SWA$ $A$ over $\Sigma$, $\Semiring$ and $\bar\Phi$,
there exists an effectively constructible \SWA
$B_{A, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$,
$B_{A, T}(t) = \displaystyle\bigoplus_{s\in \Sigma^*} A(s) \otimes T(s, t)$.
\end{proposition}
%
\begin{proof}
Let $T = \< Q, \init_T, \bar{\wei}, \final_T >$,
where $\bar{\wei}$ contains $\wei_{10}$, $\wei_{01}$, and $\wei_{11}$,
from $Q \times Q$ into respectively
$\Phi_{\Sigma}$, $\Phi_{\Delta}$, and $\Phi_{\Sigma, \Delta}$,
and let $A = \< P, \init_A, \wei_1, \final_A >$
with $\wei_1: Q \times Q \to \Phi_{\Sigma}$.
%
\noindent
The state set of $B_{A, T}$ will be $Q' = P \times Q$.
The entering, leaving and transition functions of $B_{A, T}$ will
simulate synchronized computations of $A$ and $T$,
while reading an output word of $\Delta^*$.
%
Its state entering functions is defined
for all $p \in P$, $q \in Q$
by $\init'(p, q) = \init_A(p) \otimes \init_T(q)$.
The transition function $\wei'_1$ will roughly perform
a synchronized product of transitions defined by $\wei_1$,
$\wei_{01}$ ($T$ reading in output word and not an input word)
and $\wei_{11}$ ($T$ reading both an input word and an output word).
%
Moreover, $\wei'_1$ also needs to simulate transitions
defined by $\wei_{10}$: $T$ reading in input word and not an output word.
Since $B_{A, T}$  will read only in the output word, such a transition corresponds
to an $\varepsilon$-transition of $\SWA$,
but $\SWA$ have been defined without $\varepsilon$-transitions.
Therefore, in order to take care of this case, we perform an on-the-fly
suppression of $\varepsilon$-transition in the $\SWA$ in construction,
following the algorithm of~\cite{LombardySakarovitch12ciaa}.
%
% Initialize state leaving functions i
%and $\final'(p, q) = \final_A(p) \otimes \final_T(q)$.

%Every transition of $B_{A, T}$ will
%simulate a sequence of transitions of $T$ performing the following steps:
%advance in the input word while staying immobile in the output word,
%and then make one step in the output word (and advance in the input word or not).

\noindent
Initially, for all $p_1, p_2 \in P$, and $q_1, q_2 \in Q$, let
\[
\wei'_1\bigl( \< p_1, q_1>, \< p_2, q_2>\bigr) =
\wei_1(p_1, p_2) \otimes
\bigl[
\wei_{01}(q_1, q_2)
\oplus
\displaystyle\bigoplus_{\Sigma}
\wei_{11}(q_1, q_2)
\bigr].
\]

\noindent
Iterate the following for all $p_1\in P$ and $q_1, q_2 \in Q$:
for all $p_2\in P$ and $q_3 \in Q$,
\[
\wei'_1\bigl( \< p_1, q_1>, \< p_2, q_3>\bigr) \opluseq
\displaystyle\bigoplus_{\Sigma} \wei_{10}(q_1, q_2)
\otimes
\wei'_1\bigl( \< p_1, q_2>, \< p_2, q_3>\bigr)
\]
and
\(
\final'(p_1, q_1) \opluseq
\displaystyle\bigoplus_{\Sigma} \wei_{10}(q_1, q_2)
\otimes \final'(p_1, q_2)
\)
\florent{proof correctness}
%\qed
\end{proof}

%The complexity of construction
\noindent
The construction time and size for $B_{A, T}$ are $O(\| T \|^3 . \| A \|^2)$,
where the sizes $\| T \|$ and $\| A \|$ are their number of states.
%of~$T$ is its number of states $|Q|$.
\florent{revise with nb of tr. and states}

\begin{corollary} \label{cor:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$,
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and $s \in \Sigma^+$,
there exists an effectively constructible \SWA
$B_{s, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$, $B_{s, T}(t) = T(s, t)$.
\end{corollary}

>>>>

% !TEX root = main.tex
%
% Correspondance between nested-words and trees

The hierarchical structure of nested-words, defined with the \emph{call} 
and \emph{return} markup symbols
suggest a correspondence with trees.
The lifting of this correspondence to languages, of tree automata and VPA,
has been discussed in~\cite{AlurMadhusudan09nested},
and~\cite{Caralp12VPAmult} for the weighted case.
In this section, we describe a correspondence between the symbolic-weighted extensions
of tree automata and VPA.
